An Million Transcriptome Ocular Meta-Atlas for Discovery and Development

### 2020-11-10 manuscript_00.Rmd
text updates, more figures

# meta data before doublet removal
load('~/data/massive_integrated_eye_scRNA/Mus_musculus_Macaca_fascicularis_Homo_sapiens__n_spec_genes-0__n_features2000__counts__TabulaDroplet__batch__scVI__dims8__preFilter__mindist0.001__nneighbors500.umap.Rdata')

@@

  Gene-level counts were quantified with the kallisto bustool pseudo-aligner for both the droplet and well based samples. For the droplet-based technology data, the inflection point between empty and non-empty droplets was automatically determined using the DropletUtils barcodeRank function in R. After empty droplet removal, we had `r format(cell_info_labels %>% nrow() %>% as.integer() , big.mark=",", scientific=FALSE)` cells. We then created a Seurat v3 object, calculated the percent mitochondrial genes counts, and removed cells which had more than 10% mitochondrial reads across all gene counts. For the droplet-based data, we also removed cells which had more than 3000 unique genes detected as these are likely to be doublets. After these standard quality control steps we were left with `r format(umap %>% nrow, big.mark = ",", scientific=FALSE)` cells.

@@

@@

  To rank the integration algorithms across the various benchmarks, scPOP Z scale normalizes all the metrics so that each score has an equal weight. The scores are summed which allows us to rank the performance of each integration tool across the diverse benchmarks. For users who have custom needs, the weights of each metrics can be customized to prioritize mixing or separation. scPOP finds that for our dataset scVI has the strongest performance, with a sum score of SOMETHING. The next highest performers are Harmony, fastMNN, and Scanorama. We do not claim that these results are gegnral - rather we emphasize that for each (meta) atlas creation that scPOP or some similar approach be used to identify the optimal method for each dataset.

@@

  To find the best set of parameters for integration and clustering we did a grid search across key parameters: hyper variable genes (HVG), latent dimensions, and k nearest neighbors. Using scPOP we found the optimal parameters to be 2000 HVG, 8 latent dimensions, and a resolution of 0.6 with the leiden algorithm. We also varied the UMAP projection values of nearest neighbors and minimum distance to qualitatively pick a 2D projection, selecting a minimum distance of 0.001 and 500 nearest neighbors. 
    While we removed droplet-based cells with more than 3,000 transcripts as these are highly likely to be doublets (more than one cell per droplet) there are likely many more doublets. We ran DoubletDetect and scrublet and calculated the distribution of DoubletDetect and scrublet scores across all clusters and removed clusters with a DoubletDetect and scrublet score more than 4 standard deviations above the mean. This removed another `r format((umap %>% nrow() %>% as.integer()) - (meta_filter %>% nrow() %>% as.integer()), big.mark=",", scientific=FALSE)` cells, leaving `r format((meta_filter %>% nrow()) %>% as.integer(), big.mark=",", scientific=FALSE)` in total.

## xgboost model built to label unknown cell types
  To further study cell type specific expression patterning we needed to label the `r format(umap %>% nrow() - meta_filter %>% filter(!is.na(CellType)) %>% nrow(), big.mark=",", scientific=FALSE)` unlabelled cells. Traditionally this is done by clustering the cells, then using cell type specific markers to label the clusters. However, as we had hundreds of thousands of pre-labelled cells we leveraged the expert labels by building a xgboost-based machine learning model that used the half of the labeled cells as a training set (see methods for more details) to train a cell type predictor for scEiaD input. The trained model was used to predict the cell type assignments for all cells in scEiaD. In this manner we both label most cells (cells which cannot be assigned a cell type assignment with a confidence above 0.5 were left unlabelled) and correct a small number of probable mislabels in the truth set.

@@

```{r supFig1, fig.width=5, fig.height=5, out.width=300, fig.cap=supFig2_cap, echo=FALSE, message = FALSE}
supFig2_cap <- supFig_cap(name='supFig2_cap', caption =  '')

>>>>

# meta data before doublet removal
load('~/data/massive_integrated_eye_scRNA/Mus_musculus_Macaca_fascicularis_Homo_sapiens__n_spec_genes-0__n_features2000__counts__TabulaDroplet__batch__scVI__dims8__preFilter__mindist0.001__nneighbors500.umap.Rdata')
umapRef <- umap

@@

Gene-level counts were quantified with the kallisto bustool pseudo-aligner for both the droplet and well based samples. For the droplet-based technology data, the inflection point between empty and non-empty droplets was automatically determined using the DropletUtils barcodeRank function in R. After empty droplet removal, we had `r format(cell_info_labels %>% nrow() %>% as.integer() , big.mark=",", scientific=FALSE)` cells. We then created a Seurat v3 object, calculated the percent mitochondrial genes counts, and removed cells which had more than 10% mitochondrial reads across all gene counts. For the droplet-based data, we also removed cells which had more than 3000 unique genes detected as these are likely to be doublets. After these standard quality control steps we were left with `r format(umapRef %>% nrow, big.mark = ",", scientific=FALSE)` cells.

@@

```{r figBenchmark, fig.width=6, fig.height=4, fig.cap=figBenchmark_cap, echo=FALSE, message = FALSE}
figBenchmark_cap <- fig_cap(name='figBenchmark', caption =  'a. Example of a method (Combat) which has a high level of batch blending, but poor separation of cell types (colored by cell type). b. example of an approach (no batch correction) that cleanly separates cell types but does not mix batches (colored by study). d. Optimal parameters for scVI method. sumZScale not directly to c')

source('figs_and_tables/fig_umap_benchmarking.R')
plot_grid(
  plot_grid(
    umap_plot_maker(umapCombat), 
    umap_plot_maker(umapNone, color_against = 'study_accession'),
    rel_widths = c(1.3, 1),
    labels = c('a','b')), 
  plot_grid(
    zscore_sum_all_methods, 
    zscore_droplet_scVI_optimize,
    labels = c('c','d')),
  nrow = 2,
  rel_heights = c(1, 0.4)
)
```

@@

To rank the integration algorithms across the various benchmarks, scPOP Z scale normalizes all the metrics so that each score has an equal weight. The scores are summed which allows us to rank the performance of each integration tool across the diverse benchmarks. For users who have custom needs, the weights of each metrics can be customized to prioritize batch mixing or cluster/cell type separation. scPOP finds that for our dataset scVI has the strongest performance, with a sum score of SOMETHING. The next highest performers are fastMNN, harmony, and scanorama. We do not claim that these results are gegnral - rather we emphasize that for each (meta) atlas creation that scPOP or some similar approach be used to identify the optimal method for each dataset.

@@

To find the best set of parameters for integration and clustering we did a grid search across key parameters: hyper variable genes (HVG), latent dimensions, and k nearest neighbors. Using scPOP we found the optimal parameters to be 2000 HVG, 8 latent dimensions, and a resolution of 0.6 with the leiden algorithm. We also varied the UMAP projection values of nearest neighbors and minimum distance to qualitatively pick a 2D projection, selecting a minimum distance of 0.001 and 500 nearest neighbors. 
While we removed droplet-based cells with more than 3,000 transcripts as these are highly likely to be doublets (more than one cell per droplet) there are likely many more doublets. We ran DoubletDetect and scrublet and calculated the distribution of DoubletDetect and scrublet scores across all clusters and removed clusters with a DoubletDetect and scrublet score more than 4 standard deviations above the mean. This removed another `r format((umapRef %>% nrow() %>% as.integer()) - (meta_filter %>% nrow() %>% as.integer()), big.mark=",", scientific=FALSE)` cells, leaving `r format((meta_filter %>% nrow()) %>% as.integer(), big.mark=",", scientific=FALSE)` in total.

## xgboost model built to label unknown cell types
To further study cell type specific expression patterning we needed to label the `r format(umapRef %>% nrow() - meta_filter %>% filter(!is.na(CellType)) %>% nrow(), big.mark=",", scientific=FALSE)` unlabelled cells. Traditionally this is done by clustering the cells, then using cell type specific markers to label the clusters. However, as we had hundreds of thousands of pre-labelled cells we leveraged the expert labels by building a xgboost-based machine learning model that used the half of the labeled cells as a training set (see methods for more details) to train a cell type predictor for scEiaD input. The trained model was used to predict the cell type assignments for all cells in scEiaD. In this manner we both label most cells (cells which cannot be assigned a cell type assignment with a confidence above 0.5 were left unlabelled) and correct a small number of probable mislabels in the truth set.

@@

```{r supFig1, fig.width=5, fig.height=5, out.width=300, fig.cap=supFig2_cap, echo=FALSE, message = FALSE}
supFig2_cap <- supFig_cap(name='supFig2_cap', caption =  'Performance of the various batch correction tools across various benchmarking metrics. For the LISI and Silhouette plots in A, B higher means better batch mixing and further to the right means better cluster purity. For the ARI and NMI metrics (which reflects how well cluster matches with cell type) in C, D, higher means a better score.')
